Testing gene methylation enrichment analysis approaches using simulated data

### 2023-10-27 figs/GSE158422_simulate020.Rmd
progress on simulation results

lengths <- rep(50,1000)

@@

# LA competitive mitch
simlacm <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
  # generate gene sets
  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
  lengths <- unname(unlist(lapply(genesetdatabase,length)))
  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
  # select gene sets to alter
  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
  # incorporate select changes
  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta)
  # set up limma
  mval2 <- sim$mval
  ncols <- ncol(mval2)
  groupsize <- ncols/2
  ss <- data.frame(colnames(mval2))
  colnames(ss) <- "sample"
  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
  d <- model.matrix(~ ss$case )
  dm3 <- runlimma(mval=mval2,design=d,myann=myann)
  dmagg1 <- agg(dm3,cores=4)
  m1 <- dmagg1$gme_res_df[,"mean",drop=FALSE]
  lamres1 <- runmitch(m=m1,genesets=gsets,cores=4)
  gsig_up3 <- rownames( subset( lamres1, p.adjustANOVA < 0.05 & s.dist > 0 ) )
  gsig_dn3 <- rownames( subset( lamres1, p.adjustANOVA < 0.05 & s.dist < 0 ) )
  gtup <- names(sim[[6]])
  gtdn <- names(sim[[7]])
  UPTP=length(intersect(gsig_up3 ,gtup))
  UPFP=length(setdiff(gsig_up3 ,gtup))
  UPFN=length(setdiff(gtup,gsig_up3))
  DNTP=length(intersect(gsig_dn3 ,gtdn))
  DNFP=length(setdiff(gsig_dn3 ,gtdn))
  DNFN=length(setdiff(gtdn,gsig_dn3))
  TP=UPTP+DNTP
  FP=UPFP+DNFP
  FN=UPFN+DNFN
  TN=nrow(lares1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
  PREC=TP/(TP+FP)
  REC=TP/(TP+FN)
  F1=TP/(TP+(0.5*(FP+FN)))
  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
  return(result)
}

@@

@@

num_dm_sets=50
sims=4

#groupsizes=c(3,5,8,12)
#deltas=c(0.1,0.2,0.3,0.4,0.5)
groupsizes=5
deltas=0.4

@@

>>>>

lengths <- rep(20,1000)

@@

# LA competitive mitch
simlacm <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
  # generate gene sets
  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
  lengths <- unname(unlist(lapply(genesetdatabase,length)))
  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
  # select gene sets to alter
  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
  # incorporate select changes
  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta)
  # set up limma
  mval2 <- sim$mval
  ncols <- ncol(mval2)
  groupsize <- ncols/2
  ss <- data.frame(colnames(mval2))
  colnames(ss) <- "sample"
  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
  d <- model.matrix(~ ss$case )
  dm3 <- runlimma(mval=mval2,design=d,myann=myann)
  dmagg1 <- agg(dm3,cores=4)
  m1 <- dmagg1$gme_res_df[,"mean",drop=FALSE]
  lamres1 <- runmitch(m=m1,genesets=gsets,cores=4)
  gsig_up3 <- rownames( subset( lamres1, p.adjustANOVA < 0.05 & s.dist > 0 ) )
  gsig_dn3 <- rownames( subset( lamres1, p.adjustANOVA < 0.05 & s.dist < 0 ) )
  gtup <- names(sim[[6]])
  gtdn <- names(sim[[7]])
  UPTP=length(intersect(gsig_up3 ,gtup))
  UPFP=length(setdiff(gsig_up3 ,gtup))
  UPFN=length(setdiff(gtup,gsig_up3))
  DNTP=length(intersect(gsig_dn3 ,gtdn))
  DNFP=length(setdiff(gsig_dn3 ,gtdn))
  DNFN=length(setdiff(gtdn,gsig_dn3))
  TP=UPTP+DNTP
  FP=UPFP+DNFP
  FN=UPFN+DNFN
  TN=nrow(lamres1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
  PREC=TP/(TP+FP)
  REC=TP/(TP+FN)
  F1=TP/(TP+(0.5*(FP+FN)))
  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
  return(result)
}

@@

## AA Aggregate-aggregate-limma functions

Use mean value works well here.

```{r,aafunc}

gsagg <- function(x,genesets,cores=1) {
  meds <- mclapply(1:length(genesets), function(i) {
    gs = genesets[[i]]
    xx <- x[rownames(x) %in% gs,]
    med <- apply(xx,2,mean)
  },mc.cores=cores)
  mymed <- do.call(rbind,meds)
  rownames(mymed) <- names(genesets)
  as.data.frame(mymed)
}

aalimma <- function(agag,design) {
  fit.reduced <- lmFit(agag,design)
  fit.reduced <- eBayes(fit.reduced)
  dmagg <- topTable(fit.reduced,coef=ncol(design), number = Inf)
  return(dmagg)
}

aal <- function(mval,myann,genesets,design,cores=1) {
  medf <- chragg(mval,myann,cores=cores)
  agag <- gsagg(x=medf,genesets=genesets,cores=cores)
  aalres <- aalimma(agag=agag,design=design)
  return(aalres)
}

simaa <- function(genesetdatabase, myann, mval, seed, frac_genes, frac_probes, groupsize, delta=1, num_dm_sets=50) {
  # generate gene sets
  gene_catalog <- unique(unlist(strsplit(myann$UCSC_RefGene_Name,";")))
  lengths <- unname(unlist(lapply(genesetdatabase,length)))
  gsets <- randomGeneSets(gene_catalog,lengths,seed=seed)
  # select gene sets to alter
  set.seed(seed) ; gset_mod <- sample(gsets,num_dm_sets)
  # incorporate select changes
  sim <- incorp_dm(genesets=gset_mod, myann=myann, mval=mval, seed=seed,
    frac_genes=0.5,frac_probes=0.5,groupsize=groupsize,delta=delta)
  # set up limma
  mval2 <- sim$mval
  ncols <- ncol(mval2)
  groupsize <- ncols/2
  ss <- data.frame(colnames(mval2))
  colnames(ss) <- "sample"
  ss$case <- c(rep(0,groupsize),rep(1,groupsize))
  d <- model.matrix(~ ss$case )
  # aa pipeline
  aares1 <- aal(mval=mval2, myann=myann, genesets=gsets, design=d, cores=4)
  # summarise results
  gsig_up3 <- rownames(subset(aares1, adj.P.Val < 0.05 & logFC > 0))
  gsig_dn3 <- rownames(subset(aares1, adj.P.Val < 0.05 & logFC < 0))
  gtup <- names(sim[[6]])
  gtdn <- names(sim[[7]])
  UPTP=length(intersect(gsig_up3 ,gtup))
  UPFP=length(setdiff(gsig_up3 ,gtup))
  UPFN=length(setdiff(gtup,gsig_up3))
  DNTP=length(intersect(gsig_dn3 ,gtdn))
  DNFP=length(setdiff(gsig_dn3 ,gtdn))
  DNFN=length(setdiff(gtdn,gsig_dn3))
  TP=UPTP+DNTP
  FP=UPFP+DNFP
  FN=UPFN+DNFN
  TN=nrow(aares1)-DNTP-DNFP-DNFN-UPTP-UPFP-UPFN
  PREC=TP/(TP+FP)
  REC=TP/(TP+FN)
  F1=TP/(TP+(0.5*(FP+FN)))
  result <- c("TP"=TP,"FP"=FP,"FN"=FN,"TN"=TN,"PREC"=PREC,"REC"=REC)
  return(result)
}

@@

num_dm_sets=50
sims=10

groupsizes=c(3,5,8,12)
deltas=c(0.1,0.2,0.3,0.4,0.5)
#groupsizes=5
#deltas=0.4

@@

## AA sim

```{r,runaa}

aares <- lapply(1:nrow(params) , function(j) {
  groupsize = params[j,1]
  delta = params[j,2]
  res <- mclapply(1:sims,function(i) {
    simaa(genesetdatabase=gsets, myann=myann, mval=normal_mval, seed=i*100, frac_genes=0.5, frac_probes=0.5,
      groupsize=groupsize, delta=delta, num_dm_sets=num_dm_sets)
  },mc.cores=4)
  res <- do.call(rbind,res)
  return(res)
})

aares

aares2 <- do.call(rbind,lapply(aares,colMeans))

aares2

aares2 %>% kbl(caption="AA results") %>% kable_paper("hover", full_width = F)

aares3p <- do.call(rbind,lapply(groupsizes, function (g) { aares2[params$groupsizes==g,"PREC"] }))
colnames(aares3p) <- deltas
rownames(aares3p) <- groupsizes
aares3p %>% kbl(caption="AA precision") %>% kable_paper("hover", full_width = F)

aares3r <- do.call(rbind,lapply(groupsizes, function (g) { aares2[params$groupsizes==g,"REC"] }))
colnames(aares3r) <- deltas
rownames(aares3r) <- groupsizes
aares3r %>% kbl(caption="AA recall") %>% kable_paper("hover", full_width = F)

F1(aares3p,almres3r) %>% kbl(caption="AA F1") %>% kable_paper("hover", full_width = F)
